---
title: "Reproducible Research : Week2 Assignment"
output:   md_document
---

Reproducible Research : Week2 Assignment
========================================

Set the global options for R code here:
```{r setoptions, echo=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```


## 1. Loading and preprocessing data

1.a) Let's set the working directory for the assignment.
```{r}
  setwd("~/Data_Science/Course5_ReproducibleResearch/Week2/Project")
```

1.b) Manmually download the repdata_data_activity.zip file form the link given in the assignment.
1.c) Unzip the file.
```{r}
unzip("repdata_data_activity.zip")
```

1.d) Load the file into a data frame.
```{r}
activity <- read.csv("activity.csv")
```

1.e) Clean the data by removing NAs
```{r}
activityclean <- activity[is.na(activity$steps) ==0, ]
```


## 2. Mean total number of steps taken per day
Now we will try to answer the question - What is mean total number of steps taken per day?

2.a) First, we will calculate the total number of steps taken per day.
```{r}
totstepsbyday <- aggregate(activityclean$steps, by=list(activityclean$date), FUN=sum)
colnames(totstepsbyday) <- c("date", "TotSteps")
```

2.b) Next, we will generate a Histogram of Total steps taken per day. We will use ggplot2 package for this purpose. Therefore, let's load the library.
```{r}
library(ggplot2)
```


Now that the library has been loaded, we can generate the plot.
```{r}
ggplot(totstepsbyday, aes(x=TotSteps)) + 
  geom_histogram(color="blue", fill="green", binwidth = 500) +
  labs(x="Total Steps taken by Day")
```

2.c) Lets calculate the Mean and Median of the total number of steps taken per day.
```{r}
stepsbyday_mean1 <- mean(totstepsbyday$TotSteps)
stepsbyday_median1 <- median(totstepsbyday$TotSteps)
```

***Mean for total number of steps taken per day is `r stepsbyday_mean1` and median is `r stepsbyday_median1`.***


##3. Average Daily Activity Pattern
Now we will try to answer the question - What is the average daily activity pattern?

3.a) First, we will calculate the average number of steps taken by interval (averaged across all days).
```{r}
avgstepsbyinterval <- aggregate(activityclean$steps, by=list(activityclean$interval), FUN=mean)
colnames(avgstepsbyinterval) <- c("interval", "AvgSteps")
```

3.b) Now we will create the time Series Plot.
```{r}
plot(AvgSteps~interval, avgstepsbyinterval, type="l", xlab="Time Interval", ylab="Average Steps Taken", axes=FALSE)
axis(side=1, at=c(0, 100, 200, 300, 400, 500, 600, 700, 800, 900, 1000, 1100, 1200, 1300, 1400, 1500, 1600, 1700, 1800, 1900, 2000, 2100, 2200, 2300, 2400))
axis(side=2, at=c(0, 50, 100, 150, 200, 250))
```

3.c) Get the Time Interval with the highest Average Number of Steps.
```{r}
maxstepsbyinterval <- avgstepsbyinterval[which.max(avgstepsbyinterval$AvgSteps), 1]
```

***5-minute interval containing, on average, the maximum number of steps across all the days is `r maxstepsbyinterval`.***


##4 Imputing missing values
4.a) First, we will get the total number of missing values.
```{r} 
totNA <- sum(is.na(activity$steps))
```

***Total number of missing values in the dataset is `r totNA`.***

4.b) For fixing the missing values we will replace the missing values with the mean for that 5-minute interval.

4.c) Create a new dataset that is equal to the original dataset but with the missing data filled in.

4.c.i) First, make a copy of the data frame.
```{r} 
library(data.table)
activitycomplete <- copy(activity)
```

4.c.ii) Next, find the indexes of the rows with the missing Step values.
```{r}
indxNA <- which(is.na(activity$steps), arr.ind=TRUE)
```

4.c.iii) Finally, replace the missing values with the mean of the 5-minute interval for that particular interval.
```{r}
for(i in indxNA)
{
    activitycomplete[i, 1] <-avgstepsbyinterval[avgstepsbyinterval$interval == activity[i, 3],2]      
}
```

4.d) Create a histogram and calculate mean and median of total number of steps taken per day.
4.d.i) Total number of steps taken per day with missing values imputed.
```{r}
totstepsbydayC <- aggregate(activitycomplete$steps, by=list(activitycomplete$date), FUN=sum)
colnames(totstepsbydayC) <- c("date", "TotSteps")
```

4.d.ii) Generate Histogram of Total steps taken per day.
```{r}
ggplot(totstepsbydayC, aes(x=TotSteps)) + 
  geom_histogram(color="blue", fill="green", binwidth = 500) +
  labs(x="Total Steps taken by Day", title="Total steps taken per day (missing values imputed)")
```


4.d.iii) Calculate Mean and Median
```{r}
stepsbyday_mean2 <- mean(totstepsbydayC$TotSteps)
stepsbyday_median2 <- median(totstepsbydayC$TotSteps)
```

***After imputing the missing vales, the mean for total number of steps taken per day is `r stepsbyday_mean2` and median is `r stepsbyday_median2`.***


##5 Find differences in activity patterns between weekdays and weekends

5.a) Create a new factor variable for Weekend/Weekday.

5.a.i) Add Day of the Week column.
```{r}
activitycomplete$DOW = weekdays(as.Date(activitycomplete$date))
```

5.a.ii) Add a new column specifying Weekday OR Weekend.
```{r}
activitycomplete$DayType <- ifelse(activitycomplete$DOW == "Saturday" | activitycomplete$DOW == 
                                 "Sunday", "Weekend", "Weekday")
```

5.a.iii) Convert column to factor.
```{r}
activitycomplete$DayType <- factor(activitycomplete$DayType)
```



5.b) Make a Panel plot.

5.b.i) Calculate the average number of steps taken by interval and day type (averaged across all days).
```{r}
avgstepsbyintervalC <- aggregate(steps~interval+DayType, activitycomplete, mean)
```

5.b.ii) Generate the plot.
```{r}
ggplot(avgstepsbyintervalC, aes(x=interval, y=steps)) + 
  geom_line(color="Blue") + facet_grid(DayType~.) +
  labs(x="Interval", y = "Number of Steps")
```


